### 2019-11-27 2-Chapters/1-Chapter/chapter1.tex
I fixed a lot of small typing mistakes, reported by Nathalie Mitton.
- Thanks to her!
- Shame on me haha

This third kind of wireless systems are characterized by their decentralized nature,
where communications are initialized and regulated by the devices, not by a centralized control systems.
% due to the really limited information that the base station can send to the devices (in down-link), it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
Indeed, a central control requires signaling packets that have been identified as too heavy for these kinds of systems.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common problem is the strong constraint that such IoT devices have on their power consumption, as most of them will be deployed without a direct power access and will run on a tiny battery, which lifespan should be maximized.
Most commercial IoT companies nowadays indeed sell a lifespan larger than $10$ years, like SIGFOX \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target a need for one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems (such as 4G/5G and Wi-Fi).
%
Many different standards for IoT networks have been proposed in the recent years,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layers.
To quote some examples of IoT standards, ZigBee, Z-Wave or Bluetooth are targeting short-range communications (up-to \SI{2}{\meter}), while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications (up-to \SI{50}{\kilo\meter}).
We refer to the survey \cite{Centenaro16} and references in our recent articles \cite{MoyBesson2019,MoyBesson2019Annales} for more details.

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
As this thesis focus on Reinforcement Learning models,
it is important to highlight that in such decision making models, the learner does not have access to the entire reaction of the environment after taking its action.
%
In other words, the player only sees the reward generated by playing its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of MAB in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first application of MAB since the $1930$s, with the early work of W. Thompson \cite{Thompson33},
%
as MAB are indeed a simple yet powerful example of the well-known \emph{exploration vs exploitation dilemma}.
When facing a set of $K$ actions whose effects on the environment are unknown, a learner must balance between
exploring the unknown actions, in order to collect more information about them,
and exploiting the best action, according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
The research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years \cite{bouneffouf2019survey}.

\paragraph{Opportunistic Spectrum Access.}
%
The two communities of OSA and MAB have started to interact, and the resulting works have received a great interest from both communities, since the late $2000$s and early $2010$s, with pioneers works like \cite{Liu08,Zhao10,Jouini09,Jouini10}.
The focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypotheses are maid:
the SU is equipped with a spectrum sensing capability,
and considers a fixed and finite set of $K$ orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of $K=3$ Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in time, by sub-dividing the time in discrete time steps (or rounds).
%
Thus if the SU spends a short time in the beginning of each time slot to perform sensing, it can scan for the presence or absence of any PU before transmitting, without colliding with the PU during the rest of the time slot.
If the SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be costly, in terms of energy consumption, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time.
This assumption, along with the hypothesis that PU cannot be disturbed (without which no regulation will ever be accepted for OSA), enforces the SU to transmit in the channel that it sensed, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

@@

\paragraph{MAB for OSA, and a short history of these researches at the SCEE team.}
%
Previous works of our SCEE team showed that MAB can be used to model the OSA problem:
orthogonal frequency bands (or channels) are modeled by arms $k\in\{1,\dots,K\}$,
and the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$ is modeled by a reward of $r(t) \in \{0,1\}$.
Indeed, $r(t) = 1$ indicates that no Primary User was sensed (and thus the SU can send a message), while a reward of $r(t)=0$ indicates that the channel $k$ is busy at during the time slot $t$, and no message should be sent, until the end of that slot.
%
This model was first studied by W. Jouini during his PhD thesis with C. Moy \cite{Jouini12PhD}, about ten years ago, first in the article \cite{Jouini09} and later in \cite{Jouini10,Jouini12}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned literature have essentially shown that MAB algorithms can be applied successfully on the SU side for the OSA problem.
However, if we consider CR-ready devices that cannot perform spectrum sensing, such as low-cost and low energy-consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there is no longer a distinction between PU and SU.
In other words, there is no longer any priority between users in IoT networks.
%  (no more PU, all are SU).
%
The other specificities of IoT networks can be listed as follows,
and we refer to the survey \cite{Centenaro16} for more details.
\begin{itemize}\tightlist
    % \item
    % we consider here IoT networks that run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, run on low-quality hardware, and have limited software and power capabilities (\ie, tiny batteries).
    It means that even if they are not equipped with particular spectrum sensing capabilities,
    they are equipped with a transmitter as well as with a receiver (Tx and Rx, \ie, a transceiver),
    and they have low-complexity storage and computation capabilities (\ie, small embedded processors),
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low to very low duty cycles (few messages every minute to only a few every month), and are in charge of initiating their up-link communications.
\end{itemize}

A central coordination of the devices with their base station is thus impossible as it would require them to communicate continuously with the base station, which violates the two last constraints.
%
Moreover, due to both their limited hardware (Rx/Tx), limited computational power and limited battery life, the IoT devices are able to use their received antenna to sense one channel occasionally (typically, for a few time slots after every single up-link message), but they cannot perform spectrum sensing like it is considers for OSA and CR (\ie, at the beginning of a time slot, in a ``listen-before-talk'' scheme).

@@

% \paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many MAB algorithms available, we are also interested by the question of how an engineer can choose the algorithm to implement, in a given IoT object, in order to equip this object with the capacity to adapt robustly to any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that widely used and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of CR, where many previous research works only considered the $\UCB_1$ algorithm without really justifying their choice.
This contribution was presented in the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

The second model with retransmissions concluded our collaboration with R. Bonnefoi, and resulted to a paper presented in the $1^{\text{st}}$ MoTION workshop, also during the IEEE WCNC $2019$ conference \cite{Bonnefoi2019WCNC}.

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core idea is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the entire system.
We highlight that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
This means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic about the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

@@

As explained above, the model of Chapter~\ref{chapter:4} was found intractable to analyze, mainly because we consider an IoT network with many end-devices, all following random activation patterns.
The difficulty does not reside in the fact that we are trying to analyze MAB algorithms, designed to tackle stationary problems, applied on a non-stationary problem,
but rather that we consider algorithms which are all playing in different (random) activation times, subsets of the global synchronized time steps.
Generalizing to different activation probabilities would lead to a model even harder to analyze, and enforcing at most $M \leq K$ devices is in fact not really realistic for IoT networks, even though this lead to the interesting model of Chapter~\ref{chapter:5}.

@@

% \paragraph{Chapter~\ref{chapter:6}.}
%
In order to also understand precisely how MAB algorithms behave under non-stationary environment, we studied the literature on adversarial and on non-stationary MAB, for the two cases of slowly-varying and abruptly-changing environments.
We begin the last Chapter~\ref{chapter:6} by reviewing the existing works,
and we chose to focus on the piece-wise stationary problem (\ie, abruptly-changing),
meaning that the underlying bandit problem is stationary on consecutive intervals, separated by change-points located at unknown times.
Assuming the environment to be piece-wise stationary indeed makes sense for applications to wireless networks, where a change-point can for instance corresponds to the arrival of a new group of end-devices in the network. Examples of such situation can be a new company arriving on the market in one city (\eg, the Linky counters being set-up nowadays in France), or the neighbor farmer installing sensors on his own herd of cows, etc.
%
Two main families of algorithms have been proposed for the piece-wise stationary problem,
that usually consist in combining an efficient policy designed for the stationary MAB problem (\eg, Thompson Sampling or \klUCB) and a way to adapt to changes in the arms distributions.
Passively adaptive policies use a window of a fixed or evolving size \cite{Garivier11UCBDiscount}, or a discount factor, in order to forget about the past observations \cite{Kocsis06,Gupta11thompson},
while actively adaptive policies use a statistical test to detect the change-points \cite{MellorShapiro13,Allesiardo15}.
%
As the recent literature showed that the later approach is usually more competitive, by obtaining better results from both empirical and theoretical aspects, we chose to develop our own actively adaptive algorithm.

Following two recent previous works \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, we combine an efficient index policy (\klUCB) and an efficient change-point detection test, under the assumption of bounded rewards.
We build on very recent results on the Generalized Likelihood Ratio test (GLRT) for Gaussian and sub-Gaussian variables \cite{Maillard2018GLR}, but we instead focus on bounded rewards and Bernoulli distributions.
Bounded rewards are indeed usually more appropriate for Cognitive Radio applications, as shown above.
Using the fact that bounded variables in $[0,1]$ are not only $1/4$ sub-Gaussian but also sub-Bernoulli, we prove the first finite-time guarantees for the GLRT for bounded variables.
We first show finite-time bounds on both the false alarm probability of the test, and its detection delay, under mild assumptions on the lengths of the stationary sequences.
Our algorithm is then presented in two variants, whether change-points are local (\ie, only one arm mean changes at each change-point) or global (\ie, possibly all the arm means change at a time).
%
We prove that by combining the policy \klUCB, efficient for stationary problems, and our new analysis of the GLRT for bounded rewards, we obtain state-of-the-art guarantees on the regret of the proposed algorithm, denoted \GLRklUCB.
We consider the same hypothesis as our competitors, by assuming that the length of the stationary intervals to be ``long enough'' for the change-points to be detectable.
The best regret upper-bound is obtained when the algorithm knows before-hand the horizon and the number of change-points, but our algorithm does not need to have any other knowledge of the problem difficulty to be tuned optimally.
%
The performance of \GLRklUCB{} is also illustrated on numerical experiments on synthetic data, where it is shown to outperform all the passively adaptive policies as well as the previous actively adaptive policies.
%
This last chapter is based on our last work, published at the GRETSI conference in Lille, France, in August $2019$ \cite{Besson2019Gretsi}.
This work also lead to the long version article \cite{Besson2019GLRT}, that will be partially rewritten and completed with more recent results, in order to submit it soon to a journal (probably the Journal of Machine Learning Research) before December $2019$.

@@

    % \item
\textcolor{darkgreen}{Then, the second Part~\ref{part:MABIOT}} contain three chapters (third line), that are included in both the logical and chronological orders, but can be read almost independently.
Chapter~\ref{chapter:4} starts by presenting different models of IoT networks where we show that MAB algorithms can be used with success.
Our two models are interesting and close to reality, but they appeared to be too general to propose a mathematical analysis of the good empirical performance of the considered solutions.
For this reason, we weaken the models for the rest of the document,
and both Chapters~\ref{chapter:5} and \ref{chapter:6} studies an intermediate model, lying between the stationary single-player MAB model from Chapter~\ref{chapter:2} and the intractable IoT networks models from Chapter~\ref{chapter:4}.
% \end{itemize}

@@

\item
    \emph{MALIN: ``Multi-Arm bandit Learning for Iot Networks'' with GRC: A TestBed Implementation and Demonstration that Learning Helps},
    by \textbf{Lilian Besson}, Rémi Bonnefoi, Christophe Moy.
    Demonstration presented in \emph{International Conference on Communication},
    % \footnote{~ICT 2018, see \href{http://ict-2018.org/demos}{\texttt{ict-2018.org/demos}}},
    Saint-Malo, France in June $2018$.
    See \href{https://YouTu.be/HospLNQhcMk}{\texttt{YouTu.be/HospLNQhcMk}} for a $6$-minutes presentation video.
    See Section~\ref{sec:4:gnuradio}.
    \cite{Besson2018ICT}

@@

\item
    \emph{The Generalized Likelihood Ratio Test meets klUCB: an Improved Algorithm for Piece-Wise Non-Stationary Bandits},
    by \textbf{Lilian Besson} \& Émilie Kaufmann,
    February $2019$.\\
    See Chapter~\ref{chapter:6}.
    Preprint at \href{https://HAL.Inria.fr/hal-02006471}{\texttt{HAL.Inria.fr/hal-02006471}}.
    \cite{Besson2019GLRT}\\
    An updated version, written with Odalric-Ambrym Maillard, was submitted to \href{http://www.aistats.org}{AISTATS 2020} in October $2020$.

>>>>

This third kind of wireless systems is characterized by its decentralized nature,
where communications are initialized and regulated by the devices, not by a centralized control systems.
% due to the really limited information that the base station can send to the devices (in down-link), it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
Indeed, a central control requires signaling packets that have been identified as too heavy for these kinds of systems.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common problem is the strong constraint that such IoT devices have on their power consumption, as most of them will be deployed without a direct power access and will run on a tiny battery, which lifespan should be maximized.
Most commercial IoT companies nowadays indeed sell a lifespan larger than $10$ years, like SIGFOX \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target a need for one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems (such as 4G/5G and Wi-Fi).
%
Many different standards for IoT networks have been proposed in the recent years,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layers.
To quote some examples of IoT standards, ZigBee, Z-Wave or Bluetooth are targeting short-range communications (up-to \SI{2}{\meter}), while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications (up-to \SI{50}{\kilo\meter}).
We refer to the survey \cite{Centenaro16} and references in our recent articles \cite{MoyBesson2019,MoyBesson2019Annales} for more details.

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
As this thesis focuses on Reinforcement Learning models,
it is important to highlight that in such decision making models, the learner does not have access to the entire reaction of the environment after taking its action.
%
In other words, the player only sees the reward generated by playing its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of MAB in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first application of MAB since the $1930$s, with the early work of W. Thompson \cite{Thompson33},
%
as MAB are indeed a simple yet powerful example of the well-known \emph{exploration vs exploitation dilemma}.
When facing a set of $K$ actions whose effects on the environment are unknown, a learner must balance between
exploring the unknown actions, in order to collect more information about them,
and exploiting the best action, according to its current knowledge.
%
The MAB problem has been studied in both the machine learning and the statistics communities, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
The research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years \cite{bouneffouf2019survey}.

\paragraph{Opportunistic Spectrum Access.}
%
The two communities of OSA and MAB have started to interact, and the resulting works have received a great interest from both communities, since the late $2000$s and early $2010$s, with pioneers works like \cite{Liu08,Zhao10,Jouini09,Jouini10}.
The focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypotheses are made:
the SU is equipped with a spectrum sensing capability,
and considers a fixed and finite set of $K$ orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of $K=3$ Wi-Fi channels at different frequencies, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in time, by sub-dividing the time in discrete time steps (or rounds).
%
Thus if the SU spends a short time in the beginning of each time slot to perform sensing, it can scan for the presence or absence of any PU before transmitting, without colliding with the PU during the rest of the time slot.
If the SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be costly, in terms of energy consumption, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time.
This assumption, along with the hypothesis that PU cannot be disturbed (without which no regulation will ever be accepted for OSA), enforces the SU to transmit in the channel that it sensed, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

@@

\paragraph{MAB for OSA, and a short history of these researches at the SCEE team.}
%
Previous works of our SCEE team showed that MAB can be used to model the OSA problem:
orthogonal frequency bands (or channels) are modeled by arms $k\in\{1,\dots,K\}$,
and the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$ is modeled by a reward of $r(t) \in \{0,1\}$.
Indeed, $r(t) = 1$ indicates that no Primary User was sensed (and thus the SU can send a message), while a reward of $r(t)=0$ indicates that the channel $k$ is busy during the time slot $t$, and no message should be sent, until the end of that slot.
%
This model was first studied by W. Jouini during his PhD thesis with C. Moy \cite{Jouini12PhD}, about ten years ago, first in the article \cite{Jouini09} and later in \cite{Jouini10,Jouini12}.

@@

\paragraph{Limitations and specifities of IoT networks.}
%
The aforementioned literature has essentially shown that MAB algorithms can be applied successfully on the SU side for the OSA problem.
However, if we consider CR-ready devices that cannot perform spectrum sensing, such as low-cost and low energy-consumption end-devices designed for the future Internet of Things networks, the MAB model that uses sensing to detect PU in the OSA case can no longer be applied.
Moreover, in most cases, the IoT networks use unlicensed bands, and as such there is no longer a distinction between PU and SU.
In other words, there is no longer any priority between users in IoT networks.
%  (no more PU, all are SU).
%
The other specificities of IoT networks can be listed as follows,
and we refer to the survey \cite{Centenaro16} for more details.
\begin{itemize}\tightlist
    % \item
    % we consider here IoT networks that run in unlicensed bands (no more distinction between PU and SU),
    \item
    most IoT devices are very low-cost devices, run on low-quality hardware, and have limited software and power capabilities (\ie, tiny batteries).
    It means that even if they are not equipped with particular spectrum sensing capabilities,
    they are equipped with a transmitter as well as with a receiver (Tx and Rx, \ie, a transceiver),
    and they have low-complexity storage and computation capabilities (\ie, small embedded processors),
    \item
    a single IoT base station will have to handle a very large number of devices,
    and cannot send coordination orders to them in order to optimize the network,
    \item
    IoT devices have low to very low duty cycles (few messages every minute to only a few every month), and are in charge of initiating their up-link communications.
\end{itemize}

A central coordination of the devices with their base station is thus impossible as it would require them to communicate continuously with the base station, which violates the two last constraints.
%
Moreover, due to both their limited hardware (Rx/Tx), limited computational power and limited battery life, the IoT devices are able to use their received antenna to sense one channel occasionally (typically, for a few time slots after every single up-link message), but they cannot perform spectrum sensing like it is considered for OSA and CR (\ie, at the beginning of a time slot, in a ``listen-before-talk'' scheme).

@@

% \paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many MAB algorithms available, we are also interested by the question of how an engineer can choose the algorithm to implement, in a given IoT object, in order to equip this object with the capacity to adapt robustly to any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well known existing algorithms, in Sections~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that widely used and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of CR, where many previous research works only considered the $\UCB_1$ algorithm without really justifying their choice.
This contribution was presented in the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

The second model with retransmissions concluded our collaboration with R. Bonnefoi, and resulted in a paper presented in the $1^{\text{st}}$ MoTION workshop, also during the IEEE WCNC $2019$ conference \cite{Bonnefoi2019WCNC}.

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core idea is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the entire system.
We highlight that each device communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
This means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic about the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning leads to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

@@

As explained above, the model of Chapter~\ref{chapter:4} was found intractable to analyze, mainly because we consider an IoT network with many end-devices, all following random activation patterns.
The difficulty does not reside in the fact that we are trying to analyze MAB algorithms, designed to tackle stationary problems, applied on a non-stationary problem,
but rather that we consider algorithms which are all playing in different (random) activation times, subsets of the global synchronized time steps.
Generalizing to different activation probabilities would lead to a model even harder to analyze, and enforcing at most $M \leq K$ devices is in fact not really realistic for IoT networks, even though this leads to the interesting model of Chapter~\ref{chapter:5}.

@@

% \paragraph{Chapter~\ref{chapter:6}.}
%
In order to also understand precisely how MAB algorithms behave under non-stationary environment, we studied the literature on adversarial and on non-stationary MAB, for the two cases of slowly-varying and abruptly-changing environments.
We begin the last Chapter~\ref{chapter:6} by reviewing the existing works,
and we chose to focus on the piece-wise stationary problem (\ie, abruptly-changing),
meaning that the underlying bandit problem is stationary on consecutive intervals, separated by change-points located at unknown times.
Assuming the environment to be piece-wise stationary indeed makes sense for applications to wireless networks, where a change-point can for instance corresponds to the arrival of a new group of end-devices in the network. Examples of such a situation can be a new company arriving on the market in one city (\eg, the Linky counters being set-up nowadays in France), or the neighbor farmer installing sensors on his own herd of cows, etc.
%
Two main families of algorithms have been proposed for the piece-wise stationary problem,
that usually consists in combining an efficient policy designed for the stationary MAB problem (\eg, Thompson Sampling or \klUCB) and a way to adapt to changes in the arms distributions.
Passively adaptive policies use a window of a fixed or evolving size \cite{Garivier11UCBDiscount}, or a discount factor, in order to forget about the past observations \cite{Kocsis06,Gupta11thompson},
while actively adaptive policies use a statistical test to detect the change-points \cite{MellorShapiro13,Allesiardo15}.
%
As the recent literature showed that the later approach is usually more competitive, by obtaining better results from both empirical and theoretical aspects, we chose to develop our own actively adaptive algorithm.

Following two recent previous works \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, we combine an efficient index policy (\klUCB) and an efficient change-point detection test, under the assumption of bounded rewards.
We build on very recent results on the Generalized Likelihood Ratio test (GLRT) for Gaussian and sub-Gaussian variables \cite{Maillard2018GLR}, but we instead focus on bounded rewards and Bernoulli distributions.
Bounded rewards are indeed usually more appropriate for Cognitive Radio applications, as shown above.
Using the fact that bounded variables in $[0,1]$ are not only $1/4$ sub-Gaussian but also sub-Bernoulli, we prove the first finite-time guarantees for the GLRT for bounded variables.
We first show finite-time bounds on both the false alarm probability of the test, and its detection delay, under mild assumptions on the lengths of the stationary sequences.
Our algorithm is then presented in two variants, whether change-points are local (\ie, only one arm mean changes at each change-point) or global (\ie, possibly all the arm means change at a time).
%
We prove that by combining the policy \klUCB, efficient for stationary problems, and our new analysis of the GLRT for bounded rewards, we obtain state-of-the-art guarantees on the regret of the proposed algorithm, denoted \GLRklUCB.
We consider the same hypothesis as our competitors, by assuming that the length of the stationary intervals to be ``long enough'' for the change-points to be detectable.
The best regret upper-bound is obtained when the algorithm knows before-hand the horizon and the number of change-points, but our algorithm does not need to have any other knowledge of the problem difficulty to be tuned optimally.
%
The performance of \GLRklUCB{} is also illustrated on numerical experiments on synthetic data, where it is shown to outperform all the passively adaptive policies as well as the previous actively adaptive policies.
%
This last chapter is based on our last work, published at the GRETSI conference in Lille, France, in August $2019$ \cite{Besson2019Gretsi}.
This work also leads to the long version article \cite{Besson2019GLRT}, that will be partially rewritten and completed with more recent results, in order to submit it soon to a journal (probably the Journal of Machine Learning Research) in $2020$.

@@

    % \item
\textcolor{darkgreen}{Then, the second Part~\ref{part:MABIOT}} contains three chapters (third line), that are included in both the logical and chronological orders, but can be read almost independently.
Chapter~\ref{chapter:4} starts by presenting different models of IoT networks where we show that MAB algorithms can be used with success.
Our two models are interesting and close to reality, but they appeared to be too general to propose a mathematical analysis of the good empirical performance of the considered solutions.
For this reason, we weaken the models for the rest of the document,
and both Chapters~\ref{chapter:5} and \ref{chapter:6} study an intermediate model, lying between the stationary single-player MAB model from Chapter~\ref{chapter:2} and the intractable IoT networks models from Chapter~\ref{chapter:4}.
% \end{itemize}

@@

\item
    \emph{MALIN: ``Multi-Arm bandit Learning for Iot Networks'' with GRC: A TestBed Implementation and Demonstration that Learning Helps},
    by \textbf{Lilian Besson}, Rémi Bonnefoi, Christophe Moy.
    Demonstration presented in \emph{International Conference on Telecommunications},
    % \footnote{~ICT 2018, see \href{http://ict-2018.org/demos}{\texttt{ict-2018.org/demos}}},
    Saint-Malo, France in June $2018$.
    See \href{https://YouTu.be/HospLNQhcMk}{\texttt{YouTu.be/HospLNQhcMk}} for a $6$-minutes presentation video.
    See Section~\ref{sec:4:gnuradio}.
    \cite{Besson2018ICT}

@@

\item
    \emph{The Generalized Likelihood Ratio Test meets klUCB: an Improved Algorithm for Piece-Wise Non-Stationary Bandits},
    by \textbf{Lilian Besson} \& Émilie Kaufmann,
    February $2019$.\\
    See Chapter~\ref{chapter:6}.
    Preprint at \href{https://HAL.Inria.fr/hal-02006471}{\texttt{HAL.Inria.fr/hal-02006471}}.
    \cite{Besson2019GLRT}\\
    An updated version is in writing, with Julien Seznec and Odalric-Ambrym Maillard.

### 2019-11-27 2-Chapters/2-Chapter/agregation.tex
I fixed a lot of small typing mistakes, reported by Nathalie Mitton.
- Thanks to her!
- Shame on me haha

The \Aggr{} does not have satisfying theoretical guarantees in terms of regret $R_T$, unlike many bandit algorithms.
%
Another notion, the \emph{adversarial regret}, denoted by $\overline{R_T}$,
measures the difference in terms of rewards,
between the aggregation algorithm $\Alg_{\mathrm{aggr}}$ and the best aggregated algorithm $\Alg_a$. This is in contrast with the (classical) regret, which measure the difference with the best fixed-arm strategy (Definition~\ref{def:2:regret}).
Thus, even if the aggregated algorithms have logarithmic (classical) regret, having an adversarial regret scaling as $\sqrt{T}$ does not permit to obtain a logarithmic (classical) regret for the aggregation algorithm.
%
%
Under some additional hypotheses,
\cite[Theorem 4.2]{Bubeck12} proves that
\ExpQ{} have a bounded adversarial regret, % $\overline{R_T}$
$\overline{R_T} \leq 2 \sqrt{T N \log(K)}$,
with the good choice of the learning rate sequence $(\eta_t)_{t \geq 1}$.

>>>>

The \Aggr{} does not have satisfying theoretical guarantees in terms of regret $R_T$, unlike many bandit algorithms.
%
Another notion, the \emph{adversarial regret}, denoted by $\overline{R_T}$,
measures the difference in terms of rewards,
between the aggregation algorithm $\Alg_{\mathrm{aggr}}$ and the best aggregated algorithm $\Alg_a$. This is in contrast with the (classical) regret, which measure the difference with the best fixed-arm strategy (Definition~\ref{def:2:regret}).
Thus, even if the aggregated algorithms have logarithmic (classical) regret, having an adversarial regret scaling as $\sqrt{T}$ does not permit to obtain a logarithmic (classical) regret for the aggregation algorithm.
%
%
Under some additional hypotheses,
\cite[Theorem 4.2]{Bubeck12} proves that
\ExpQ{} has a bounded adversarial regret, % $\overline{R_T}$
$\overline{R_T} \leq 2 \sqrt{T N \log(K)}$,
with the good choice of the learning rate sequence $(\eta_t)_{t \geq 1}$.
